Computing Direct Sum Decompositions

### 2025-04-23 paper/main.tex
changed proof of graded case

Let $R$ be a $\N$-graded ring, with $R_0 = k$, an algebraically closed field.
%Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
%%the effective cone
%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a degree zero endomorphism $\phi \in [\End M]_0$.

@@

\begin{lem}\label{lem:graded}
  Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
\end{lem}
\begin{proof}
  Fix homogeneous generators $m_1,\dots,m_N$ of $M$, with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$. The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
  \[ \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_n
  \end{pmatrix}, \]
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$. In particular, $B_i$ is the induced endomorphism on the submodule generated by all generators of degree $d_i$.

  Since the matrices $B_i$ have entries in the algebraically closed field $k$, by performing row and column operations over $k$ we can put each $B_i$ into Jordan canonical form $J_i$ with entries in $k$. In particular, since the generators of degree $d_i$ are linearly independent, a $k$-linear change of basis on this set induces an automorphism $\alpha_i$ of $M$.

  Let $\alpha$ be the composition of these automorphisms for each block $B_i$, then $\phi' = \alpha^{-1}\circ\phi\circ\alpha$ corresponds to performing row and column operations on $\phi$ over $k$ and has the form
  \begin{equation}
    \label{eq:Jordan}
    \phi' = \begin{pmatrix}
      J_1    & *      & \dots  & *      \\
      0      & J_2    & \dots  & *      \\
      \vdots & \vdots & \ddots & \vdots \\
      0      & 0      & \dots  & J_n
    \end{pmatrix},
  \end{equation}
  where each $J_i$ is a Jordan block over $k$ with eigenvalue $\lambda_i$, and the remaining entries are in $R$.
  Let $\mu_i$ be the multiplicity of $\lambda_i$ and let $\psi_i = (\phi' - \lambda_i \id_M)^{\mu_i}$.
  \mahrud{TODO}

  %% Relabeling, we may assume that $\lambda_i\neq 0$ for $i=1,\dots,n_0$ and $\lambda_i=0$ for $i=n_0+1,\dots,n$.

  %% If we take $N_0$ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{n_0+1},\dots,J_n$, then for $N\geq N_0$ we have
  %% $$
  %% \phi^N =
  %% \begin{pmatrix}
  %%   J_1^N & *** &\dots & \dots& \dots  &\dots & *** \\
  %%   0 & J_2^N & \dots &\dots&\dots & \dots& *** \\
  %%   \vdots &\vdots & \ddots &\vdots \\
  %%   0 & 0 & \dots & J_{n_0}^N & *** & \dots & ***\\
  %%   0 & 0 & \dots & 0 & 0 & \dots & 0\\
  %%   \vdots &\vdots & \ddots &\vdots \\
  %%   0 & 0 & \dots & 0 & 0 & \dots & 0
  %% \end{pmatrix},
  %% $$
  %% where each nonzero $J_i^N$ has the units $\lambda_i^N$ on the diagonal.
  %% ($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)

  %% Fix $N$ as above.
  %% Now, consider the map $ \phi^N: \im(\phi^N) \to \im(\phi^N)$. If this map is surjective, then $\im(\phi^N)$ is a direct summand of $M$, with complementary summand $\coker \phi^N$: to see this, observe that the injection $\im(\phi^N) \inc M$ would then  compose with $\phi^N$ to be a surjection, and since $\im(\phi^N)$ is a finitely generated module, a surjection must be an isomorphism $\a:\im \phi^N \to \im \phi^N$ by \cite[Corollary~4.4]{Eisenbud95}. Thus, we obtain a splitting of the inclusion $\im \phi^N \inc M$ via $\a\inv\circ \phi^N$; since $\a\inv$ is an isomorphism, we have that $\im \phi^N$ is (isomorphic to) a direct summand of $M$, with complementary summand $\coker \phi^N$.

  %% So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
  %% To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\lambda_{i'}\in k$).
  %% That is, a generating set is given by
  %% $$u_1 e_1,\quad  u_2 e_2+a_{21}e_1,\quad \dots ,\quad u_{n-n_0}e_{n-N_0}+a_{n-N_0,n-1}e_{n-N_0-1}+\dots + a_{N-n_0,1}e_1.$$
  %% Since the $u_i$ are units, the $R$-span of these generators is clearly the same as the $R$-span of $e_1,\dots,e_{n-N_0}$, and thus after applying $\psi$ we get back the $R$-span of $f_1,\dots,f_{n-N_0} $, which is $\im \psi^N$. Thus, $\phi: \im \psi^N \to \im \psi^N$ is surjective, and the result follows.
\end{proof}

@@

% The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$. The following corollary of the proof of \Cref{lem:graded} ensures that we can always assume $\phi$ has zero as an eigenvalue:

%% \begin{lem}
%%   The eigenvalues of $\phi$ lie in $k$. In particular, for any $\phi \in [\End M]_0$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in [\End M]_0$ has zero as an eigenvalue.
%% \end{lem}

By ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - \lambda \id_M)$\mahrud{define determinant over a module, or say determinant of its cover?}, which is a univariate polynomial in $\lambda$ with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.

%% \begin{proof}
%%   The expression of $\phi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - \lambda \id_M$ is the product of $\lambda_i - \lambda$, thus the eigenvalues are the $\lambda_i$, which all lie in $k$.
%% \end{proof}

>>>>

Let $R$ be a $\N$-graded ring, with $R_0 = k$ a field.
%Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
%%the effective cone
%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a general degree zero endomorphism $\phi \in [\End M]_0$.

@@

\begin{lem}\label{lem:graded}
  \mahrud{rewrite statement} Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
\end{lem}
\begin{proof}
  Fix homogeneous generators $m_1,\dots,m_N$ of $M$, with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$. The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
  \begin{equation}\label{eq:blocks}
    \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_n
    \end{pmatrix},
  \end{equation}
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$. In particular, $B_i$ is the induced endomorphism on the submodule generated by all generators of degree $d_i$.

  Note that the characteristic polynomial $\det(\phi - \lambda \id_M) = \prod \det(B_i - \lambda \id)$ is a univariate polynomial in $\lambda$ with coefficients in $k$ which may have solutions in the algebraic closure of $k$. We will refer to the collection of $k$-eigenvalues of $B_i$ as the ``eigenvalues of $\phi$''.

  By generality of $\phi$ suppose it has eigenvalues $\lambda_1,\dots,\lambda_r$ in $k$. Let $\mu(\lambda_j)$ be the maximum multiplicity of $\lambda$ in each diagonal block and set $\psi_j = (\phi - \lambda_j \id_M)^{\mu(\lambda_j)}$. We will show that $\psi_j$ is a split surjection, hence $M = \im\psi_j \oplus \coker\psi_j$. Furthermore, note that if $\phi_j$ and $\phi_k$ are different then $M = \coker\psi_j \oplus \coker\psi_k \oplus \im(\psi_j\circ\psi_k)$, thus we have a decomposition
  \begin{equation}\label{eq:decomposition}
    M = \coker\psi_1 \oplus \cdots \oplus \coker\psi_r \oplus \im\psi.
  \end{equation}

  To see that each $\psi_j$ is a split surjection, note that the set of generators of $M$ with degree $d_i$ are linearly independent and we can view each $B_i$ as a $k$-linear transformation of the $k$-vector space spanned by these generators. Thus, performing $k$-linear row and column operations on $(B_i - \lambda_j \id)^{\mu(\lambda_j)}$, we may write it as a diagonal matrix with ones and zeros on the diagonal. In particular, this is a $k$-linear change of basis on the degree $d_i$ generators of $M$ and induces an automorphism $\alpha_{i,j}$ of $M$.

  Let $\alpha$ be the composition of these automorphisms for each $B_i$ and $\lambda_j$, then $\alpha^{-1}\circ\psi_j\circ\alpha$ corresponds to performing $k$-linear row and column operations and has ones and zeros on the diagonal. Finally, performing $R$-linear row and column operations, we may clear the rows and columns with a one on the diagonal to get $\psi'_j$, whose image and cokernel are summands and are isomorphic to the image and cokernel of $\psi_j$.
\end{proof}

@@
